---
title: "Modelling _E. coli_ die-off in mini trials"
author: "Nicholas Waters"
date: "2/5/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Introduction

## The Data

### Main dataset
Data is stored in this repo's `data` dir.  The `raw` directory contains the raw data files recieved, which were copied to the `clean` dir, and modified as needed. 

The bulk of the data was from file "Box Behnken analysis.xlsx", recieved from RN via Skype on 2019-02-27, and this file was an updated version of the earlier "All Raw Data Mini Trial.xlsx" file.  The data were in the "Block on rep" and "methane" tabs.

The main changes to "Box Behnken analysis" that were made:
- label OLR was changed to OL to reflect the time-independant nature of the value
- label HRT was changed to time, for the inverse reason
- the methane aggretation was fixed to prevent values from being summed twice
- addition of an "export" tab, for exporting to csv.

All these changes can be found in "./data/clean/Box Behnken analysis_nick_27.02.19.xlsx", and exported as "./data/clean/trial_data.csv".

### Day 0 Values
After discussion about the experimental setup and the move to a time-based model rather than the Box Behnken analysis, we requested the Day 0 values for the 3 feedstocks, innoculum, and their mixtures. The file "Mini Trial FS FIB TS&VS.xlsx" was recieved 2019-04-30 from SN.  

Data was aggregated and exported as "./data/clean/day0.csv" from the "day0" worksheet I added.


### Questions about the data

- When is the data refereing to the Feedstock, Innoculum, Food Waste, Recipe, etc?
- do we have TS for the different conditions at day 0?
- Have I aggregated the data properly?
- Should indicator counts include innoculum?

## Data explorations

First thing we did was read in the main data file and merge it with the day 0 data.
```{r reading_data}


```
