---
title: "Modelling _E. coli_ die-off in mini trials"
author: "Nicholas Waters"
date: "2/5/2019"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(rstan)
library(GGally)

options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)
```

## Introduction

## The Data

### Main dataset
Data is stored in this repo's `data` dir.  The `raw` directory contains the raw data files recieved, which were copied to the `clean` dir, and modified as needed. 

The bulk of the data was from file "Box Behnken analysis.xlsx", recieved from RN via Skype on 2019-02-27, and this file was an updated version of the earlier "All Raw Data Mini Trial.xlsx" file.  The data were in the "Block on rep" and "methane" tabs.

The main changes to "Box Behnken analysis" that were made:
- label OLR was changed to OL to reflect the time-independant nature of the value
- label HRT was changed to time, for the inverse reason
- the methane aggretation was fixed to prevent values from being summed twice
- addition of an "export" tab, for exporting to csv.

All these changes can be found in "./data/clean/Box Behnken analysis_nick_27.02.19.xlsx", and exported as "./data/clean/trial_data.csv".

### Day 0 Values
After discussion about the experimental setup and the move to a time-based model rather than the Box Behnken analysis, we requested the Day 0 values for the 3 feedstocks, innoculum, and their mixtures. The file "Mini Trial FS FIB TS&VS.xlsx" was recieved 2019-04-30 from SN.  

Data was aggregated and exported as "./data/clean/day0.csv" from the "day0" worksheet I added.

### Processed data
The results of the two datasets are combined and written to the `clean` folder as `cleaned_combined_minitrial_data.csv`.

### A note about experimental design and terms
Three things are mixed in these reractors: slurry, Food waste, and the iccoculum (from an AD reactor)
Recipe may be used interchangably with Feedstock.
There are 3 recipes, with different slurry:food waste ratios of 1:2, 1:3, and 3:1.

- *OL*: organic load, expressed in g VS/L
- sCOD: the soluble Chemical Oxygen Demand


## Todo
- incorporate innoculumn sCOD values when recieved; either retest orr impute pH.  

## Data explorations

First thing we did was read in the main data file and merge it with the day 0 data. Then, we convert log-scale the the indicator bacteria counts, adding a dummy value of 1 to prevent errors of logging 0.


```{r reading_data}
sn_raw <- rbind(
  read.csv("./data/clean/trial_data.csv"),
  read.csv("./data/clean/day0.csv", skip=3, skipNul = T)
)
sn <- sn_raw %>%
  rowwise() %>% 
  transform(
    Recipe=as.factor(Recipe),
    Ecoli =  log10(Ecoli + 1),
    Coliforms = log10(Coliforms + 1),
    Enterococci = log10(Enterococci + 1))
summary(sn)

write.csv(sn, "data/clean/cleaned_combined_minitrial_data.csv")
```

Next, we can do some explortatory analysis:

```{r}
ggscatmat(sn,  color="Recipe", alpha = .3) + theme(legend.position = "bottom")
#ggpairs(sn, aes(color=Recipe, alpha=.5))

```

Initial observations:
- the original Box Behnken design can be seen looking at the top left area, seeing the distribution of values between Time, OL, and Temp.
- E. coli and Coliforms are, unsurprisingly, very highly correlated.
- TS and VS are highly correlated


Lets replot without some of the highly correlated ones that

```{r}
ggscatmat(sn %>% select(-TS, -Coliforms),  color="Recipe", alpha = .3) + theme(legend.position = "bottom")
```

sCOD still seems high day 0 in FWS13, but that my be biologically relavent.


```{r}
ggplot(sn, aes(x=VS, y=TS)) + geom_point() + facet_grid(~OL)  + geom_smooth()
ggplot(sn, aes(x=VS, y=TS)) + geom_point() + facet_grid(~Temp) + geom_smooth()
ggplot(sn, aes(x=VS, y=TS, color=Ammonia)) + geom_point() + facet_grid(~Temp) + geom_smooth()
ggplot(sn, aes(x=VS, y=TS, color=Ammonia)) + geom_point() + facet_grid(~Temp + Recipe) + geom_smooth()
ggplot(sn, aes(x=Time)) + geom_histogram() + facet_grid(~OL + Time)
```


## Modelling

```{r}
stan_data = list(
    nbugs=3,
    nrows=nrow(sn),
    Temp=(sn$Temp -19)/18 + 1,
    Ammonia=sn$Ammonia,
    pH=sn$pH,
    Ecoli=sn$Ecoli,
    Coliforms=sn$Coliforms,
    Time=sn$Time,
    OL=sn$OL,
    Enterococci=sn$Enterococci,
    sCOD=sn$sCOD,
    VS=sn$VS,
    TS=sn$TS,
    Recipe=as.numeric(sn$Recipe),
    Methane=sn$Methane
  )
fit <- stan(
  file = "./models/model_min.stan",
  verbose = T,
  chains=4,
  data = stan_data
)

ggplot(data.frame(x = c(-5, 5)), aes(x)) +
    stat_function(fun = dcauchy, n = 1e3, args = list(location = 1, scale = 1))
stan_dens(fit, pars=c("recipe_effect") )
stan_dens(fit, pars=c("load_effect"))
plot(fit, pars=c("recipe_effect") )
plot(fit, pars=c("load_effect"))
plot(fit, pars=c("temp_effect"))
plot(fit, pars=c("lp__"))
plot(fit, pars=c("sigma"))
plot(fit, chains=T)

stan_model(fit)
mle = optimizing(stan_model("./models/model_min.stan"), data=stan_data)
```



# change log

- 2019-05-02: after meeting, update calculations for day 0 values. 

- 2019-05-01: setup .Rmd, incorporate inital day 0 values
