---
title: "Cleaning minitrial data"
author: "Nicholas Waters"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```
## change log

- 2019-09-04: removed bayesian modelling, updated plot colors, removed extraneous information.
- 2019-06-04: improve model fit and description
- 2019-05-03: fixed sCOD values, fixed methane values in export
- 2019-05-02: after meeting, update calculations for day 0 values. 
- 2019-05-01: setup .Rmd, incorporate inital day 0 values

## Introduction

## The Data

### Main dataset
Data is stored in this repo's `data` dir.  The `raw` directory contains the raw data files recieved, which were copied to the `clean` dir, and modified as needed. 

The bulk of the data was from file "Box Behnken analysis.xlsx", recieved from RN via Skype on 2019-02-27, and this file was an updated version of the earlier "All Raw Data Mini Trial.xlsx" file.  The data were in the "Block on rep" and "methane" tabs.

The main changes to "Box Behnken analysis" that were made:

- label OLR was changed to OL to reflect the time-independant nature of the value
- label HRT was changed to time, for the inverse reason
- the methane aggretation was fixed to prevent values from being summed twice
- addition of an "export" tab, for exporting to csv.

All these changes can be found in "./data/clean/Box Behnken analysis_nick_27.02.19.xlsx", and exported as "./data/clean/trial_data.csv".

### Day 0 Values
After discussion about the experimental setup and the move to a time-based model rather than the Box Behnken analysis, we requested the Day 0 values for the 3 feedstocks, innoculum, and their mixtures. The file "Mini Trial FS FIB TS&VS.xlsx" was recieved 2019-04-30 from SN.  These were not used for PS's modelling, but can be ncluded for some visualizations.

Data was aggregated and exported as "./data/clean/day0.csv" from the "day0" worksheet I added.

### Processed data
The results of the two datasets are combined and written to the `clean` folder as `cleaned_combined_minitrial_data.csv`.

### A note about experimental design and terms
Three things are mixed in these reactors: slurry, food waste, and the iccoculum (from an AD reactor)
Recipe may be used interchangably with Feedstock.
There are 3 recipes, with different slurry:food waste ratios of 1:2, 1:3, and 3:1.

- *OL*: organic load, expressed in g VS/L
- sCOD: the soluble Chemical Oxygen Demand



## Data explorations

First thing we did was read in the main data file and merge it with the day 0 data. Because the day zero are true for all temperatures, we duplicate the data for the 19 and 55 degree conditions. Then, we convert log-scale the the indicator bacteria counts, adding a dummy value of 1 to prevent errors of logging 0.


```{r reading_data}
main_data <- read.csv("./data/clean/trial_data.csv")
day0 <- read.csv("./data/clean/day0.csv", skip=3, skipNul = T, stringsAsFactors = F)
day0 <- day0 %>% select(-fs_conc, -fs_vol, -inoc_conc, -inoc_vol) %>% spread(key = measurement, value = Value)
day0<- rbind(
  day0,
  day0 %>% transform(Temp=replace(Temp,Temp==37, 19)),
  day0 %>% transform(Temp=replace(Temp,Temp==37, 55))
) 
names(main_data)
names(day0)
sn_raw <- rbind(main_data, day0)

sn <- sn_raw %>%
  rowwise() %>% 
  transform(
    Recipe=as.factor(Recipe),
    Ecoli =  log10(Ecoli + 1),
    Coliforms = log10(Coliforms + 1),
    Enterococci = log10(Enterococci + 1)) %>%
  transform(
    Ecoli =  ifelse(Ecoli < 1, 1, Ecoli), 
    Coliforms = ifelse(Coliforms < 1, 1, Coliforms), 
    Enterococci =  ifelse(Enterococci < 1, 1, Enterococci))

summary(sn)
```

It may prove useful to note the components of the three recipes.  The organic loading rates were calcualted based on the volitile solid quantity.  the recepies are ratios of 2 of three ingrrerdiates (not including the innoculum, which is added as discussed above):  dairy cattle slurry(DCS), Fats Oils and Grease (FOG), and bakery waste(BW).

- FWS12: FOG:DCS 1:2
- FWS13: DCS:BW 1:3 
- FWS31: DCS:BW 3:1  

We add those proportion to the data.  Then, we save the cleaned, prepated data. 

```{r}
#name, FOG, BW, DCS
(recipes <- as.data.frame(
  matrix(dimnames = list(c("FWS12", "FWS13", "FWS31"), c( "FOG", "BW", "DCS", "init_Coliforms", "init_Ecoli", "init_Enterococci")), c(
           1/3, 0 , 2/3, 6.87, 6.48, 5.49,
           0, 1/4,3/4, 6.91, 6.77, 6.2,
           0, 3/4,1/4, 6.4, 6.33, 4.93), ncol=6, byrow = T)))
recipes$Recipe <- rownames(recipes)
sn <- merge(sn, recipes, by="Recipe")
write.csv(sn, "data/clean/cleaned_combined_minitrial_data.csv")
```
